\textbf{SPECFEM3D}\\ \textbf{Cartesian}\\ \textbf{User Manual}

### 2022-11-09 doc/USER_MANUAL/01_introduction.tex
Path updates for parameters moved from constants.h to Par_file and vice versa, and other minor corrections.

To use coupling with FK, just use the set of parameters that is in the DATA/Par\_file input file of the code: \\
%
\begin{verbatim}
#-----------------------------------------------------------
#
# Coupling with an injection technique (DSM, AxiSEM, or FK)
#
#-----------------------------------------------------------
COUPLE_WITH_INJECTION_TECHNIQUE = .false.
INJECTION_TECHNIQUE_TYPE        = 3       # 1 = DSM, 2 = AxiSEM, 3 = FK
MESH_A_CHUNK_OF_THE_EARTH       = .false.
TRACTION_PATH                   = ./DATA/AxiSEM_tractions/3/
FKMODEL_FILE                    = FKmodel
RECIPROCITY_AND_KH_INTEGRAL     = .false. # does not work yet
\end{verbatim}
That part (coupling with FK) is actively maintained and works fine. See e.g. \href{https://doi.org/10.1002/2014GL061644}{GRL\_Ping\_Tong\_2014} for some examples.
There is also an example that is provided with the code:\\
\texttt{specfem3d/EXAMPLES/small\_example\_coupling\_FK\_specfem}. \\

@@

\item {\bf Gravity}: SPECFEM3D can now perform gravity field calculations in addition (or instead of) seismic wave propagation only.
See flag \texttt{GRAVITY\_INTEGRALS} in file \texttt{setup/constants.h.in}. Please also refer to \href{https://doi.org/10.1093/gji/ggx010}{GJI\_Martin\_gravimetry\_2017}.
And yes, that is the reason why Dimitri added a falling apple on the cover of the manual :-).
Note that SPECFEM3D can also model transient gravity perturbations induced by earthquake rupture, as developed and explained in \cite{HaAmBaChMoSoWh2015}.
These are two different things, and both are implemented and avaible in SPECFEM3D.
To use the second feature, please refer to \texttt{doc/how\_to/README\_gravity\_perturbation.txt}.\\

>>>>

To use coupling with FK, just use the set of parameters that is in the \texttt{DATA/Par\_file} input file of the code: \\
%
\begin{verbatim}
#-----------------------------------------------------------
#
# Coupling with an injection technique (DSM, AxiSEM, or FK)
#
#-----------------------------------------------------------
COUPLE_WITH_INJECTION_TECHNIQUE = .false.
INJECTION_TECHNIQUE_TYPE        = 3       # 1 = DSM, 2 = AxiSEM, 3 = FK
MESH_A_CHUNK_OF_THE_EARTH       = .false.
TRACTION_PATH                   = ./DATA/AxiSEM_tractions/3/
FKMODEL_FILE                    = FKmodel
RECIPROCITY_AND_KH_INTEGRAL     = .false. # does not work yet
\end{verbatim}
That part (coupling with FK) is actively maintained and works fine. See e.g. \href{https://doi.org/10.1002/2014GL061644}{GRL\_Ping\_Tong\_2014} for some examples.
There is also an example that is provided with the code:\\
\texttt{specfem3d/EXAMPLES/small\_example\_coupling\_FK\_specfem}. \\

@@

\item {\bf Gravity}: SPECFEM3D can now perform gravity field calculations in addition (or instead of) seismic wave propagation only.
See flag \texttt{GRAVITY\_INTEGRALS} in file \texttt{setup/constants.h}. Please also refer to \href{https://doi.org/10.1093/gji/ggx010}{GJI\_Martin\_gravimetry\_2017}.
And yes, that is the reason why Dimitri added a falling apple on the cover of the manual :-).
Note that SPECFEM3D can also model transient gravity perturbations induced by earthquake rupture, as developed and explained in \cite{HaAmBaChMoSoWh2015}.
These are two different things, and both are implemented and avaible in SPECFEM3D.
To use the second feature, please refer to \texttt{doc/how\_to/README\_gravity\_perturbation.txt}.\\

### 2022-11-09 doc/USER_MANUAL/02_getting_started.tex
Path updates for parameters moved from constants.h to Par_file and vice versa, and other minor corrections.

The SPECFEM3D Cartesian software package relies on the SCOTCH library
to partition meshes created with CUBIT. METIS \citep{KaKu98a,KaKu98c,KaKu98b}
can also be used instead of SCOTCH if you prefer, by changing the parameter
\texttt{PARTITIONING\_TYPE} in the \texttt{Par\_file}. You will also then need to install
and compile Metis version 4.0 (do {*}NOT{*} install Metis version
5.0, which has incompatible function calls) and edit \texttt{Makefile.in}
and uncomment the METIS link flag in that file before running \texttt{configure}.\\

@@

Regular POSIX I/O can be problematic when dealing with large simulations on large
clusters (typically more than $10,000$ MPI processes). SPECFEM3D can use the ADIOS library~\cite{Liu2013}
to take advantage of advanced parallel file system features. To enable
ADIOS, the following steps should be done:
\begin{enumerate}
\item Install ADIOS (available from \url{https://www.olcf.ornl.gov/center-projects/adios/}).
Make sure that your environment variables reference it.
\item You may want to change ADIOS related values in the \texttt{constants.h.in} file.
The default values probably suit most cases.
\item Configure using the \texttt{-{}-with-adios} flag.
\end{enumerate}
ADIOS is currently only usable for meshfem3D generated mesh (i.e. not for meshes generated
with CUBIT). Additional control parameters are discussed in section~\ref{cha:Main-Parameter}.

@@

Now that you have set the compiler information, you need to select
a number of flags in the \texttt{constants.h} file depending on your
system:
\begin{description}
\item [{\texttt{LOCAL\_PATH\_IS\_ALSO\_GLOBAL}}] Set to \texttt{.false.}
on most cluster applications. For reasons of speed, the (parallel)
distributed database generator typically writes a (parallel) database
for the solver on the local disks of the compute nodes. Some systems
have no local disks, e.g., BlueGene or the Earth Simulator, and other
systems have a fast parallel file system, in which case this flag
should be set to \texttt{.true.}. Note that this flag is not used
by the database generator or the solver; it is only used for some
of the post-processing.
\end{description}
%
The package can run either in single or in double precision mode.
The default is single precision because for almost all calculations
performed using the spectral-element method using single precision
is sufficient and gives the same results (i.e. the same seismograms);
and the single precision code is faster and requires exactly half
as much memory. Select your preference by selecting the appropriate
setting in the \texttt{constants.h} file:
\begin{description}
\item [{\texttt{CUSTOM\_REAL}}] Set to \texttt{SIZE\_REAL} for single precision
and \texttt{SIZE\_DOUBLE} for double precision.
\end{description}
%
In the \texttt{precision.h} file:
\begin{description}
\item [{\texttt{CUSTOM\_MPI\_TYPE}}] Set to \texttt{MPI\_REAL} for single
precision and \texttt{MPI\_DOUBLE\_PRECISION} for double precision.
\end{description}
%
On many current processors (e.g., Intel, AMD, IBM Power), single precision
calculations are significantly faster; the difference can typically
be 10\% to 25\%. It is therefore better to use single precision. What
you can do once for the physical problem you want to study is run
the same calculation in single precision and in double precision on
your system and compare the seismograms. If they are identical (and
in most cases they will), you can select single precision for your
future runs.\\

>>>>

The SPECFEM3D Cartesian software package relies on the SCOTCH library
to partition meshes created with CUBIT. METIS \citep{KaKu98a,KaKu98c,KaKu98b}
can also be used instead of SCOTCH if you prefer, by changing the parameter
\texttt{PARTITIONING\_TYPE} in the \texttt{DATA/Par\_file}. You will also then need to install
and compile Metis version 4.0 (do {*}NOT{*} install Metis version
5.0, which has incompatible function calls) and edit \texttt{Makefile.in}
and uncomment the METIS link flag in that file before running \texttt{configure}.\\

@@

Regular POSIX I/O can be problematic when dealing with large simulations on large
clusters (typically more than $10,000$ MPI processes). SPECFEM3D can use the ADIOS library~\cite{Liu2013}
to take advantage of advanced parallel file system features. To enable
ADIOS, the following steps should be done:
\begin{enumerate}
\item Install ADIOS (available from \url{https://www.olcf.ornl.gov/center-projects/adios/}).
Make sure that your environment variables reference it.
\item You may want to change ADIOS related values in the \texttt{setup/constants.h} file.
The default values probably suit most cases.
\item Configure using the \texttt{-{}-with-adios} flag.
\end{enumerate}
ADIOS is currently only usable for meshfem3D generated mesh (i.e. not for meshes generated
with CUBIT). Additional control parameters are discussed in section~\ref{cha:Main-Parameter}.

@@

Now that you have set the compiler information, you need to select
a number of flags in the \texttt{setup/constants.h} file depending on your
system:
\begin{description}
\item [{\texttt{LOCAL\_PATH\_IS\_ALSO\_GLOBAL}}] Set to \texttt{.false.}
on most cluster applications. For reasons of speed, the (parallel)
distributed database generator typically writes a (parallel) database
for the solver on the local disks of the compute nodes. Some systems
have no local disks, e.g., BlueGene or the Earth Simulator, and other
systems have a fast parallel file system, in which case this flag
should be set to \texttt{.true.}. Note that this flag is not used
by the database generator or the solver; it is only used for some
of the post-processing.
\end{description}
%
The package can run either in single or in double precision mode.
The default is single precision because for almost all calculations
performed using the spectral-element method using single precision
is sufficient and gives the same results (i.e. the same seismograms);
and the single precision code is faster and requires exactly half
as much memory. Select your preference by selecting the appropriate
setting in the \texttt{setup/constants.h} file:
\begin{description}
\item [{\texttt{CUSTOM\_REAL}}] Set to \texttt{SIZE\_REAL} for single precision
and \texttt{SIZE\_DOUBLE} for double precision.
\end{description}
%
In the \texttt{precision.h} file:
\begin{description}
\item [{\texttt{CUSTOM\_MPI\_TYPE}}] Set to \texttt{MPI\_REAL} for single
precision and \texttt{MPI\_DOUBLE\_PRECISION} for double precision.
\end{description}
%
On many current processors (e.g., Intel, AMD, IBM Power), single precision
calculations are significantly faster; the difference can typically
be 10\% to 25\%. It is therefore better to use single precision. What
you can do once for the physical problem you want to study is run
the same calculation in single precision and in double precision on
your system and compare the seismograms. If they are identical (and
in most cases they will), you can select single precision for your
future runs.\\

### 2022-11-09 doc/USER_MANUAL/03_mesh_generation.tex
Path updates for parameters moved from constants.h to Par_file and vice versa, and other minor corrections.

Please note that your Vp- and Vs-speeds are given for a reference frequency. To change
this reference frequency, you change the value of \texttt{ATTENUATION\_f0\_REFERENCE}
in the main constants file \texttt{constants.h} found in subdirectory
\texttt{src/shared/}.
The code uses a constant $Q$ quality factor, write(IMAIN,*) "but approximated based on a series of Zener standard linear solids (SLS).
The approximation is thus performed in a given frequency band determined based on that \texttt{ATTENUATION\_f0\_REFERENCE} reference frequency.
\item [{anisotropic\_flag}] Flag describing the anisotropic model to use
in case an anisotropic simulation should be conducted. See the file
\texttt{model\_aniso.f90} in subdirectory \texttt{src/generate\_databases/}
for an implementation of the anisotropic models. In case no anisotropy
is available, it can be set to zero.
\end{description}

@@

where \texttt{\textbf{domain\_ID}}\textbf{ is 1 for acoustic and 2
for elastic or viscoelastic materials,} \texttt{material\_ID} a unique
identifier, \texttt{rho} the density in $kg\, m^{-3}$, \texttt{vp}
the P-wave speed in $m\, s^{-1}$, \texttt{vs} the S-wave speed in
$m\, s^{-1}$, \texttt{Q} the quality factor and \texttt{anisotropy\_flag}
an identifier for anisotropic models.
Note that both \texttt{Qkappa} and \texttt{Qmu} are ignored
by the code unless \texttt{ATTENUATION} is set. If you want a model
with no \texttt{Qmu} attenuation, both set \texttt{ATTENUATION} to
\texttt{.false.} in the \texttt{Par\_file} and set \texttt{Qmu} to
9999 here. If you want a model with no \texttt{Qkappa} attenuation, set \texttt{Qkappa} to 9999 here.
Note that Qmu is always equal to Qs, but Qkappa is in general not equal to Qp.

@@

The partitioning is done in serial for now (in the next release we
will provide a parallel version of that code). It needs to be run in the main directory because it expects the \texttt{./DATA/Par\_file}.
The synopsis is:
%
\begin{verbatim}
./bin/xdecompose_mesh nparts input_directory output_directory
\end{verbatim}
%
\noindent
where
\begin{itemize}
\item \texttt{nparts} is the number of partitions, i.e., the number of cores
for the parallel simulations,
\item \texttt{input\_directory} is the directory which holds all the files
generated by the Python script \texttt{cubit2specfem3d.py} explained
in the previous Section~\ref{subsec:Exporting-the-Mesh}, e.g. \texttt{./MESH/},
and
\item \texttt{output\_directory} is the directory for the output of this
partitioner which stores ACII-format files named like \texttt{proc{*}{*}{*}{*}{*}{*}\_Database}
for each partition. These files will be needed for creating the distributed
databases, and have to reside in the directory \texttt{LOCAL\_PATH}
specified in the main \texttt{Par\_file}, e.g. in directory \texttt{./OUTPUT\_FILES/DATABASES\_MPI}.
Please see Chapter~\ref{cha:Creating-Distributed-Databases} for
further details.
\end{itemize}
Note that all the files generated by the Python script \texttt{cubit2specfem3d.py}
must be placed in the \texttt{input\_directory} folder before running
the program.

@@

by \texttt{NGLLX} in the \texttt{constants.h} file. We generally use

@@

Must be set at least to \texttt{1} if \texttt{USE\_REGULAR\_MESH} is set to \texttt{.true.}.

@@

(only interpreted if \texttt{USE\_REGULAR\_MESH} is set to \texttt{.true.}).

@@

layer (only interpreted if \texttt{USE\_REGULAR\_MESH} is set to \texttt{.true.}

@@

The mesher generates the necessary partitions in parallel, one set
for each of the $\nprocxi\times\nproceta$ slices that constitutes
the mesh (see Figure~\ref{fig:For-parallel-computing}). After the
mesher finishes, you can log in to one of the compute nodes and view
the contents of the \texttt{LOCAL\_PATH} directory to see the files
generated by the mesher. These files will be needed for creating the
distributed databases, and have to reside in the directory \texttt{LOCAL\_PATH}
specified in the main \texttt{Par\_file}, e.g. in directory \texttt{OUTPUT\_FILES/DATABASES\_MPI}.
Please see Chapter~\ref{cha:Creating-Distributed-Databases} for
further details.
%
\item [{\texttt{NMATERIALS}}] The number of different materials in your
model. In the following lines, each material needs to be defined as:
%
\begin{itemize}[leftmargin=-8ex,itemindent=8ex]
\item[] acoustic/elastic material
{\small
\begin{verbatim}
material_ID rho vp vs Qkappa Qmu anisotropy_flag domain_ID
\end{verbatim}
}
\item[] poroelastic material
{\small
\begin{verbatim}
material_id rho_s rho_f phi tort kxx kxy kxz kyy kyz kzz kappa_s kappa_f kappa_fr eta mu_fr domain_id
\end{verbatim}
}
\item[] tomographic
{\small
\begin{verbatim}
material_id type-keyword domain-name tomo-filename tomo_id domain_id
\end{verbatim}
}
\end{itemize}
For acoustic and elastic material, you would specify
\begin{itemize}
\item \texttt{material\_ID} : a material ID number (1,2,..) to identify the material for the following mesh region selection.
\item \texttt{rho}, \texttt{vp}, \texttt{vs} : density (in $kg/m^3$), $V_p$ and $V_s$ (in $m/s$) wave speeds
\item \texttt{Qkappa, Qmu} : quality factor (0=no attenuation) for bulk $Q_{\kappa}$ and shear attenuation $Q_{\mu}$
\item \texttt{anisotropy\_flag} : 0=no anisotropy / 1,2,.. check with implementation
in \texttt{aniso\_model.f90}
\item \texttt{domain\_id} : 1=acoustic / 2=elastic / 3=poroelastic
\end{itemize}
\noindent
For poroelastic materials, you would specify the corresponding poroelastic properties.
Please read the section below for more on poroelastic parameters.\\

@@

Please note that the mesher suggests
a time step \texttt{DT} to run the solver with. The mesher output
file also contains a table about the quality of the mesh to indicate
possible problems with the distortions of elements. Alternatively,
output can be directed to the screen instead by uncommenting a line
in \texttt{constants.h}:
{\small
\begin{verbatim}
! uncomment this to write messages to the screen
! integer, parameter :: IMAIN = ISTANDARD_OUTPUT
\end{verbatim}}

>>>>

Please note that your Vp- and Vs-speeds are given for a reference frequency. To change
this reference frequency, you change the value of \texttt{ATTENUATION\_f0\_REFERENCE}
defined (in Hz) in input file \texttt{DATA/Par\_file}.
The code uses a constant $Q$ quality factor, write(IMAIN,*) "but approximated based on a series of Zener standard linear solids (SLS).
The approximation is thus performed in a given frequency band determined based on that \texttt{ATTENUATION\_f0\_REFERENCE} reference frequency.
\item [{anisotropic\_flag}] Flag describing the anisotropic model to use
in case an anisotropic simulation should be conducted. See the file
\texttt{model\_aniso.f90} in subdirectory \texttt{src/generate\_databases/}
for an implementation of the anisotropic models. In case no anisotropy
is available, it can be set to zero.
\end{description}

@@

where \texttt{\textbf{domain\_ID}}\textbf{ is 1 for acoustic and 2
for elastic or viscoelastic materials,} \texttt{material\_ID} a unique
identifier, \texttt{rho} the density in $kg\, m^{-3}$, \texttt{vp}
the P-wave speed in $m\, s^{-1}$, \texttt{vs} the S-wave speed in
$m\, s^{-1}$, \texttt{Q} the quality factor and \texttt{anisotropy\_flag}
an identifier for anisotropic models.
Note that both \texttt{Qkappa} and \texttt{Qmu} are ignored
by the code unless \texttt{ATTENUATION} is set. If you want a model
with no \texttt{Qmu} attenuation, both set \texttt{ATTENUATION} to
\texttt{.false.} in the \texttt{DATA/Par\_file} and set \texttt{Qmu} to
9999 here. If you want a model with no \texttt{Qkappa} attenuation, set \texttt{Qkappa} to 9999 here.
Note that Qmu is always equal to Qs, but Qkappa is in general not equal to Qp.

@@

The partitioning is done in serial for now (in the next release we
will provide a parallel version of that code). It needs to be run in the main directory because it expects the \texttt{DATA/Par\_file}.
The synopsis is:
%
\begin{verbatim}
./bin/xdecompose_mesh nparts input_directory output_directory
\end{verbatim}
%
\noindent
where
\begin{itemize}
\item \texttt{nparts} is the number of partitions, i.e., the number of cores
for the parallel simulations,
\item \texttt{input\_directory} is the directory which holds all the files
generated by the Python script \texttt{cubit2specfem3d.py} explained
in the previous Section~\ref{subsec:Exporting-the-Mesh}, e.g. \texttt{./MESH/},
and
\item \texttt{output\_directory} is the directory for the output of this
partitioner which stores ACII-format files named like \texttt{proc{*}{*}{*}{*}{*}{*}\_Database}
for each partition. These files will be needed for creating the distributed
databases, and have to reside in the directory \texttt{LOCAL\_PATH}
specified in the main \texttt{DATA/Par\_file}, e.g. in directory \texttt{./OUTPUT\_FILES/DATABASES\_MPI}.
Please see Chapter~\ref{cha:Creating-Distributed-Databases} for
further details.
\end{itemize}
Note that all the files generated by the Python script \texttt{cubit2specfem3d.py}
must be placed in the \texttt{input\_directory} folder before running
the program.

@@

by \texttt{NGLLX} in the \texttt{setup/constants.h} file. We generally use

@@

Must be set at least to \texttt{1} if \texttt{USE\_REGULAR\_MESH} is set to \texttt{.false.}.

@@

(only interpreted if \texttt{USE\_REGULAR\_MESH} is set to \texttt{.false.}).

@@

layer (only interpreted if \texttt{USE\_REGULAR\_MESH} is set to \texttt{.false.}

@@

The mesher generates the necessary partitions in parallel, one set
for each of the $\nprocxi\times\nproceta$ slices that constitutes
the mesh (see Figure~\ref{fig:For-parallel-computing}). After the
mesher finishes, you can log in to one of the compute nodes and view
the contents of the \texttt{LOCAL\_PATH} directory to see the files
generated by the mesher. These files will be needed for creating the
distributed databases, and have to reside in the directory \texttt{LOCAL\_PATH}
specified in the main \texttt{DATA/Par\_file}, e.g. in directory \texttt{OUTPUT\_FILES/DATABASES\_MPI}.
Please see Chapter~\ref{cha:Creating-Distributed-Databases} for
further details.
%
\item [{\texttt{NMATERIALS}}] The number of different materials in your
model. In the following lines, each material needs to be defined as:
%
\begin{itemize}[leftmargin=-8ex,itemindent=8ex]
\item[] acoustic/elastic material
{\small
\begin{verbatim}
material_ID rho vp vs Qkappa Qmu anisotropy_flag domain_ID
\end{verbatim}
}
\item[] poroelastic material
{\small
\begin{verbatim}
material_id rho_s rho_f phi tort kxx kxy kxz kyy kyz kzz kappa_s kappa_f kappa_fr eta mu_fr domain_id
\end{verbatim}
}
\item[] tomographic
{\small
\begin{verbatim}
material_id type-keyword domain-name tomo-filename tomo_id domain_id
\end{verbatim}
}
\end{itemize}
For acoustic and elastic material, you would specify
\begin{itemize}
\item \texttt{material\_ID} : a material ID number (1,2,..) to identify the material for the following mesh region selection.
\item \texttt{rho}, \texttt{vp}, \texttt{vs} : density (in $kg/m^3$), $V_p$ and $V_s$ (in $m/s$) wave speeds
\item \texttt{Qkappa, Qmu} : quality factor (0=no attenuation) for bulk $Q_{\kappa}$ and shear attenuation $Q_{\mu}$
\item \texttt{anisotropy\_flag} : 0=no anisotropy / 1,2,.. check with implementation
in \texttt{aniso\_model.f90}
\item \texttt{domain\_id} : 1=acoustic / 2=elastic / 3=poroelastic
\end{itemize}
\noindent
For poroelastic materials, you would specify the corresponding poroelastic properties.
Please read the section below for more on poroelastic parameters.\\

@@

Please note that the mesher suggests
a time step \texttt{DT} to run the solver with. The mesher output
file also contains a table about the quality of the mesh to indicate
possible problems with the distortions of elements. Alternatively,
output can be directed to the screen instead by uncommenting a line
in \texttt{setup/constants.h}:
{\small
\begin{verbatim}
! uncomment this to write messages to the screen
! integer, parameter :: IMAIN = ISTANDARD_OUTPUT
\end{verbatim}}
